---
title: "API-Shiny Pipeline"
author: "Leah Jackman"
date: "4/13/2020"
output: pdf_document
---

# Learning Objectives

At the end of this lesson, you should be able to...

* Gain insight into the role of a data scientist in industry 

* Interact with simple APIs.
  + Find and read API documentation.
  + Submit a GET request for data.
  + Parse data from the API.

* Create simple Shiny applications.
  + Construct a UI function.
    + Understand the basics of a DOM (Document Object Model).
    + Understand the basic R functions that construct the DOM. 
  + Construct a Server function.
    + Understand observe functions.
    + Understand render functions.
    + Understand reactive variables.

# The Real World
Part of what a Data Scientist does is...

Find data, assess data, analyze data, visualize data

# The Real World: Find Data
**Boss:** Find the latitude and longitude of all the Walgreens and Boots locations world-wide. When can you get this done?

**Me:** (Never worked with APIs, maps, geospatial analysis) ...

**Boss:** When can you get this done?

**Me:** A week?

# The Real World: Assess Data
**Me:** There's an indicator column in this database, but all the values are 0. (Knows there must be at least one 1.) Any idea why?

**Database Owner/IT Chatbot:** Negative. See. Owner. Of. Table.

**Owner of Table/Overworked Project Manager:** (shrug)

**Me:** (Guess I can't trust the data in that column. I wonder about the rest of the table...)

# The Real World: Analyze Data
**Old School Industry Statistician:** Here is a file with thousands of comments from doctors. Use NLP to find all the negative ones.

**Me:** (Confident, having just read Tidy Text Mining) I can do a sentiment analysis on the comments, shouldn't take me too long!

**Me:** (Proud, returns with data) Here!

**OSIS:** Now use NLP to tell me why the doctors are upset.

**Me:** (Unsure) Well I can do some clustering to see which comments are related and analyze to see if the clusters are meaningful...

**OSIS:** No, I want quotes from the actual comments that state the exact reason the doctors are upset.

**Me:** (Looking at data - no labels, just free text)... I don't think NLP can do that ...

**OSIS:** (Frown)

# The Real World: Visualize Data
**Me:** (Demos fancy new data visualization)

**Boss:** This isn't sexy enough. I saw an application that has a spinny-globe loading screen, can you add that?

**Me:** Well... Maybe... I can try ...But what does that have to do with this visualization?

**Boss:** SPINNY GLOBE! SEXY! BRING ME!

# APIs

* API stands for Application Programming Interface. 
* Serves as a mediator/gatekeeper. 
* Takes as input a request in a very strict format.
* Request is parsed, and initiates some set of processes. 
* Processes return data objects in a structured format - JSON, XML, etc. 

# Requests
[httr package](https://cran.r-project.org/web/packages/httr/vignettes/quickstart.html)

[jsonlite package](https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html)

# GET

**Me:** (Goes to API's front door. Piles of data with my name on it sitting there.) Thanks API!  

**API:** You're welcome, use the data wisely or you will be sued. Have a lovely day!

# GET Example

```{r, eval = FALSE}
#This request returns the state data aggregated over all time.

#Libraries
library(httr)
library(jsonlite)

#https://documenter.getpostman.com/view/8854915/SzS8rjHv?version=latest State Level Data
#set a GET request
covid_states_req <- GET("https://covidtracking.com/api/states", query = list("state" = "NY"))

status_code <- covid_states_req$status_code
    
#"good" request when status = 200. 
if (status_code == 200) {
  #get content of request. 
  #many ways to do this, but I like the following because the output is a data frame
  data_df <- jsonlite::fromJSON(content(covid_states_req, as = "text"))
  
  #"good" means correct format of request, but might still be issue with query itself
  #if issue with query, return error message
  if ("error" %in% names(data_df)) {
    error_msg <- data_df$message
    data_df <- NULL
    
  }else {
    data_df <- data_df
  }
}

head(data_df)

```

**You Try!**
Write a request that returns all data, not aggregated data, for a given state. 
Hint: [Read the documentation](https://documenter.getpostman.com/view/8854915/SzS8rjHv?version=latest)

# POST

**Me:** (Goes to API's front door, door closed) Knock, knock. Here's my req-

**API:** (Interrupting) Intruder! Intruder!

**Me:** Oh, right right. (Puts eyeball up to optical scanner, completes secret knock knock pattern)

**API:** Phew! OK, now what do you want?

**Me:** (Sends list of exacting instructions.)

**API:** (Scrutinizing) Looks to be in order. Here's your data. Use it wisely or you will be sued. Have a lovely day!


Including an example here, but we're not going to go over it.

# POST Example

```{r, eval = FALSE}
#' Search Citeline for Protocol Information
#' 
#' Searches Citeline Trial endpoint based on user input 
#' 
#' @param return_fields_lst a list of API fields to return
#' 
#' @param search_fields_lst a list of API fields to filter on
#' 
#' @param operators_lst a list of API operators to apply to search_field_lst and search_values_lst
#' 
#' @param search_values_lst a list values used to filter search_field_lst
#' 
#' @param user_name_txt the user name for the search. Should be standard CI credentials.
#' 
#' @param password_txt the password for the search. Should be standard CI credentials.
#' 
#' @return The first 100 matches in a dataframe. One column per return_field_lst value.
#' 
#' @export

search_similar_trials_fn_gbl <- function(return_fields_lst, 
                                         search_fields_lst,
                                         operators_lst,
                                         search_values_lst,
                                         user_name_txt,
                                         password_txt,
                                         session){
  library(httr)
  
  url <- "https://identity.pharmaintelligence.informa.com/connect/token"
  auth_token <- 
    "Y3VzdG9tZXJfYXBpX2NsaWVudF9uYXRpdmU6NTIwNzhjMGItMTI5Mi00MGVhLWFkYjAtOWE4MWY4OGNjZDMy"
  authHeader <- add_headers(Authorization = paste("Basic", auth_token))
  
  post_req <- POST(url,
                   authHeader,
                   body = list(
                     grant_type = "password",
                     scope = "customer-api",
                     username = user_name_txt,
                     password = password_txt))
  post_output <- content(post_req, "parsed")
  token <- post_output$access_token
  
  api_url <- "https://api.pharmaintelligence.informa.com"
  endpoint <- paste0("/v1/search/trial")
  usage_url <- paste0(api_url, endpoint)
  
  authHeader <- add_headers(Authorization = paste("Bearer", token),
                            Accept = "application/json")
  
  #Note: Do not follow this example! This is before I learned about jsonlite::toJSON. 
  #All you need to do is create a list of lists and use toJSON
  #ex) json_req <- list(and = list(field = "investigator_name", value = "Dr. Person")) %>% toJSON()
  start_query <- '\"and\":['
  primary_query <- paste0(paste0('{\"', operators_lst, '\": {\"value\": \"', search_values_lst, 
                                 '\",\"name\": \"', search_fields_lst, '\"}}'), collapse = ",")
  end_query <- ']'
  
  query_list <- paste0(start_query, primary_query, end_query)
  
  query <- paste0("{\"fields\": [\"", paste0(return_fields_lst, collapse = "\",\""), "\"],", query_list, "}")
  
  post_req <- POST(usage_url, authHeader, body = query, content_type_json())
  get_output <- httr::content(post_req, "parsed")
  
  # Error Handling based on get_output's status code
  if (get_output$meta$statusCode == 400) {
    error_title <- "Bad Request"
    error_message <- "The request contains invalid parameters, or invalid or incomplete data"
  } else if (get_output$meta$statusCode == 401) {
    error_title <- "Unauthorized"
    error_message <- "You failed to provide an authentication header, or your credentials are invalid or expired"
  } else if (get_output$meta$statusCode == 403) {
    error_title <- "Forbidden"
    error_message <- "You tried to perform an action that is not authorised. E.g. you failed to request the required scope"
  } else if (get_output$meta$statusCode == 404) {
    error_title <- "Not Found"
    error_message <- "The resource or endpoint does not exist. This could be temporary or permanent"
  } else if (get_output$meta$statusCode == 406) {
    error_title <- "Not Acceptable"
    error_message <- "The API does not support the requested format in the Accept header. Try again using one of the support formats"
  } else if (get_output$meta$statusCode == 500) {
    error_title <- "Internal Server Error"
    error_message <- "You should never receive this error because our clever developers catch them all... but if you are unlucky enough to get one, please report it to us at pharmaapifeedback@informa.com"
  } else if (get_output$meta$statusCode == 503) {
    error_title <- "Service Unavailable"
    error_message <- "The API has been switched to maintenance mode for a major or complex release"
  }
  
  # Display error message
  if (get_output$meta$statusCode != 200)
    sendSweetAlert(session = session, title = error_title, text = error_message, type = "error")
  
  # Dealing with pagination
  if (length(get_output$items) > 0) {
    # Names of output fields
    output_items <- get_output$items
     next_page <- get_output$pagination$nextPage

     while(!is.null(next_page)){
       get_next_req <- GET(next_page, authHeader)
       get_next_output <- httr::content(get_next_req, "parsed")
       output_items <- append(output_items, get_next_output$items)
       next_page <- get_next_output$pagination$nextPage
     }
    return(output_items)
  } else {
    return(NA)
  }
  
}
```

# API Tips and Tricks

* Read the documentation!
* Use the verbose() option to troubleshoot.
* Special Characters - either escape them with \\ or use I()


# Shiny

[Shiny](https://shiny.rstudio.com/) is a package in R that constructs websites. 

Two ways to construct a website in Shiny

* ui.R, server.R
* ui function, server function together in one .R file.


# UI

* Creates the original DOM (Document Object Model)
* Outlines what objects (sidebars, navigation menus, boxes, inputs, etc.) exist, and where they exist in a website.

Demo "inspect" on website to show example.

# UI Example
[Simple Example from RStudio](https://shiny.rstudio.com/gallery/kmeans-example.html)
Uses the two separate .R files approach

```{r, eval = FALSE}

vars <- setdiff(names(iris), "Species")

pageWithSidebar(
  headerPanel('Iris k-means clustering'),
  sidebarPanel(
    #each input has an associated class
    selectInput('xcol', 'X Variable', vars),
    selectInput('ycol', 'Y Variable', vars, selected = vars[[2]]),
    numericInput('clusters', 'Cluster count', 3, min = 1, max = 9)
  ),
  mainPanel(
    #each output has an associated class
    plotOutput('plot1')
  )
)

```

**You Try!**
What other types of inputs exist beside select and numeric?


# Server

* Tells the website how it should behave based on interactivity.
* For the most part, hidden from user, so "Inspect" not as helpful.
* Input variables from the ui can be accessed using "input$_input id_"
* Input variables from the ui cannot be accessed, but can be defined using "output$_output id_"

# Server Example
[Simple Example from RStudio](https://shiny.rstudio.com/gallery/kmeans-example.html)
Uses the two separate .R files approach

```{r, eval = FALSE}

function(input, output, session) {

  # Combine the selected variables into a new data frame
  # selectedData is available to every function in the server,
  #and will automatically change if xcol or ycol changes
  # note the ({}) syntax - this is use for all reactivity
  selectedData <- reactive({
    #inputs accessed here
    iris[, c(input$xcol, input$ycol)]
  })

  #instead of reactive, can also use observe(), observeEvent() depending on needs
  clusters <- reactive({
    kmeans(selectedData(), input$clusters)
  })

  #setting the output object
  output$plot1 <- renderPlot({
    palette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
      "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999"))

    par(mar = c(5.1, 4.1, 0, 1))
    plot(selectedData(),
         col = clusters()$cluster,
         pch = 20, cex = 3)
    points(clusters()$centers, pch = 4, cex = 4, lwd = 4)
  })

}

```

**You Try!**
Many packages other than ggplot2 exist to create visualizations in R. Find one such package How would you amend the above code to use the new package? 

# COVID-19 Example

[GitHub Project](https://github.com/ldjack5/teaching)

Download and play around.
Some things to try:

* Make the graph prettier.
* Add a select input for model types - Exponential, Logistic, etc. - that would be appropriate for this data. When the user selects a model, overlay the model on the graph.
* Check out shinyDashboard and shinyDashboardPlus and try to add a theme to make the website more appealing.
* Add a checkbox with "confirmed" and "deaths". Use the checkbox to allow the user to see either confirmed, or deaths or both on the graph.
* Change the country select input to a selectizeInput where you can accept multiple values. Plot the curves for all countries on a single graph.
* Add in states/provinces for countries other than the US.


# Reflection

At the end of this lesson, you should be able to...

* Gain insight into the role of a data scientist in industry 

* Interact with simple APIs.
  + Find and read API documentation.
  + Submit a GET request for data.
  + Parse data from the API.

* Create simple Shiny applications.
  + Construct a UI function.
    + Understand the basics of a DOM (Document Object Model).
    + Understand the basic R functions that construct the DOM. 
  + Construct a Server function.
    + Understand observe functions.
    + Understand render functions.
    + Understand reactive variables.
